Rhyme: A Query Language for Nested Data Structures

### 2024-12-13 paper.tex
update section 2.1
finish section 2.2

In this section, we will walk through the end-to-end workflow of \rhyme, demonstrating how the \rhyme query get compiled and generate efficient code to execute the query. Figure \ref{workflow} demonstrates the pipeline of \rhyme compiler.\par

@@

      \label{workflow}

@@

\rhyme provides multiple frontends, including the Pipe API, Textual API, and JavaScript API. As a domain-specific language (DSL) embedded in JavaScript, \rhyme allows queries to be represented as JSON objects. This design enables users to construct queries directly in JSON format. For instance, to group the sum of values in a dataset by its keys, users can specify \inline{data..key} as the object query's key and \inline{data..value} as the value, as shown in the example below. Notably, because \inline{sum(data..value)} is nested within \inline{data..key}, the logic variable in \inline{sum(data.*.value)} is bound, making the operation a group-by sum.

@@

To support code generation for strongly typed languages like C/C++, \rhyme allows users to optionally provide type schemas for the input data. As illustrated in Figure \ref{typeschema}, users can specify, for example, that the data consists of objects with 8-bit unsigned integer (u8) keys and string values. This type schema is essential for backend code generation, as it provides precise type information for each sub-expression of the query, enabling the generation of highly specialized and efficient code.
\begin{figure}[H]
  \centering
      \includegraphics[scale=0.4]{figures/typeschema.png}
      \caption{Example type schema of \rhyme}
      \label{typeschema}
  \end{figure}
\subsection{\rhyme Middle-end}
\subsection{\rhyme Backend}
\section{Type System}\label{typesystem}
\section{Sparse Tensor Algebra}\label{sparse}
\section{Evaluation}\label{evaluation}
\section{Conclusion}\label{conclusion}
%\newpage
\bibliographystyle{acm}
\bibliography{paper}
\end{document}
\endinput

>>>>

In this section, we will walk through the end-to-end workflow of \rhyme, demonstrating how the \rhyme query get compiled and generate efficient code to execute the query. Figure \ref{workflowfig} demonstrates the pipeline of \rhyme compiler.\par

@@

      \label{workflowfig}

@@

\rhyme provides multiple frontends, including the Pipe API, Textual API, and JavaScript API. As a domain-specific language (DSL) embedded in JavaScript, \rhyme allows queries to be represented as JSON objects. This design enables users to construct queries directly in JSON format. For instance, to group the sum of values in a dataset by its keys, users can specify \inline{data.*.key} as the object query's key and \inline{data.*.value} as the value, as shown in the example below. Notably, because \inline{sum(data.*.value)} is nested within \inline{data.*.key}, the logic variable in \inline{sum(data.*.value)} is bound, making the operation a group-by sum.

@@

To support code generation for strongly typed languages like C/C++, \rhyme allows users to optionally provide type schemas for the input data. As illustrated in Figure \ref{typeschema}, users can specify, for example, that the data consists of objects with 8-bit unsigned integer (u8) keys and string values. This type schema is essential for backend code generation, as it provides precise type information for each sub-expression of the query, enabling the generation of highly specialized and efficient code. We will talk about the type system more extensively in Section \ref{typesystem}.
      \includegraphics[scale=0.4]{figures/typeschema2.png}
      \caption{Example type schema of \rhyme}
      \label{typeschema}
\subsection{\rhyme Middle-end}
After the query is parsed into AST and the type schema is provided, we will perform type checking and inference and generate the intermediate representation. As shown in Figure \ref{typeschema}, we can infer the type of sub-expressions based on the given input type schema. If the input data has type \inline{{u8 : string}}, we can infer that the key of data \inline{*} (i.e. the logic variable) has type u8, we can also infer that \inline{data.*} has type string. We will perform both type inference and type checking to make sure that the type system conforms and we can correctly type all sub-expressions.\par
The \rhyme AST is a structured representation of the query, but we need a more fine-grained representation of the program that can capture both the data and control dependencies.
\rhyme has two types of IRs: \emph{generators} and \emph{assignments}. \emph{generators} are instructions used for iterating over a collection of data. Logic variables are responsible for iterating over data and directly map to generators. For example, \inline{data.*} corresponds to a generator that iterate over \inline{data}. \emph{assignments} are instructions used for computing the final and intermediate results. For example, \inline{sum(data.*)} corresponds to a sum operation that aggregate the value of \inline{data}.\rhyme also utilizes temporary variables to store materialized intermediate results.\par
It is worth noting that the IR does not enforce strict program structures. The program Structure is implied by the dependencies between statements. When an assignment uses a generator, it will cause a generator-assignment dependency (i.e. control dependency). When an assignment uses the result of an intermediate expression, it will cause a assignment-assignment dependency (i.e. data dependency). The \rhyme compiler will utilize these dependencies to produce an optimal program structure that respect the original query's semantic. Due to the dependencies-based IR design, \rhyme can perform multiple optimization like loop hoisting and code motion very easily.\par
After the IR generation, we will also assign types to each IR nodes. We will record the type of each generator and intermediate temporary variables. These type signature will be necessary when we generate code in the backend. We will also record for each generator the type of data sources they are iterating over. And we will generate specialized efficient code to co-iterate multiple data sources for one generator. For example, if one generator is co-iterating over multiple sparse vectors, we can generate specialized code to co-advancing the sparse vectors for efficiency as sparse vector does not offer constant-time look-up. We will cover this more in detail in Section \ref{sparse}.
After the query is parsed into an AST and the type schema is provided, \rhyme performs type checking and inference, followed by the generation of an intermediate representation (IR). As illustrated in Figure \ref{typeschema}, the type of sub-expressions can be inferred based on the input type schema. For example, if the input data is typed as \inline{\{u8 : string\}}, we can infer that the key of \inline{data} (i.e., the logic variable \inline{*}) has type \inline{u8}, while \inline{data.*} itself has type \inline{string}. Both type inference and type checking ensure that the type system is conformed and that all sub-expressions are correctly typed.

While the \rhyme AST provides a structured representation of the query, it is insufficient for capturing both data and control dependencies. For this purpose, \rhyme introduces two types of IRs: generators and assignments. Generators are instructions for iterating over data collections, with logic variables directly mapping to these generators. For example, \inline{data.*} corresponds to a generator that iterates over \inline{data}. Assignments, on the other hand, represent computations of intermediate and final results. For instance, \inline{sum(data.*)} corresponds to an aggregation operation that computes the sum of \inline{data} values. Temporary variables are also used to store materialized intermediate results.

The \rhyme IR does not enforce a rigid program structure; instead, the structure is implied by the dependencies between statements. A generator used by an assignment introduces a generator-assignment dependency (control dependency), while one assignment using the result of another creates an assignment-assignment dependency (data dependency). These dependencies guide the \rhyme compiler in generating an optimal program structure that preserves the original query semantics. This dependency-based IR design facilitates advanced optimizations such as loop hoisting and code motion.

Following IR generation, types are assigned to each IR node. The compiler records the types of generators and temporary variables, as well as the types of data sources associated with each generator. This type information is critical for backend code generation, enabling specialized and efficient co-iteration of multiple data sources. For example, if a generator co-iterates over multiple sparse vectors, \rhyme generates optimized code for advancing the sparse vectors efficiently, accounting for their lack of constant-time look-up. A more detailed explanation of these optimizations is provided in Section \ref{sparse}.
\subsection{\rhyme Backend}
\section{Type System}\label{typesystem}
\section{Sparse Tensor Algebra}\label{sparse}
\section{Evaluation}\label{evaluation}
\section{Conclusion}\label{conclusion}
%\newpage
\bibliographystyle{acm}
\bibliography{paper}
\end{document}
\endinput
